Lensing in the Schwarzschild Spacetime Applicable to Neutron Stars

### 2023-10-04 src/tex/ms.tex
calculated the momenta

The metric has two killing vectors, $\partial_t$ and $\partial_\phi$, since it
is independent of the corresponding coordinates.  Write
\begin{equation}
\partial_t = \xi^\mu \partial_\mu, \qquad \partial_\phi = \zeta^\mu \partial_\mu;
\end{equation}
then the only non-zero component of $\xi^\mu$ is $\xi^t = 1$, and the only
non-zero component of $\zeta^\mu$ is $\zeta^\phi = 1$.  If $p^\mu$ is the
momentum associated with a geodesic,
\begin{equation}
p^\mu = \frac{\dd \chi^\mu(\lambda)}{\dd \lambda},
\end{equation}
with $\lambda$ either proper time (for time-like geodesics) or an affine
parameter (null geodesics), and $\chi^\mu(\lambda)$ the coordinates of the
points on the geodesic, then N\"{o}ther's Theorem gives 
\begin{eqnarray}
\label{eq:conserved-e}
g_{\mu \nu} \xi^\mu p^\nu = - \left( 1 - \frac{2}{\chi^r} \right) p^t & = & \textnormal{const} \equiv - e,\\
\label{eq:conserved-l}
g_{\mu \nu} \zeta^\mu p^\nu = (\chi^r)^2 \sin^2 (\chi^\theta) p^\phi & = & \textnormal{const} \equiv l
\end{eqnarray}
along any geodesic.  ($e$ and $l$ are the energy and angular momentum---per unit
mass, if appropriate---of the particle traveling the geodesic measured by an
observer at infinity.)

For a null geodesic, we also have the normalization condition 
\begin{equation}
\label{normalization}
g_{\mu \nu} p^\mu p^\nu = 0.
\end{equation}
Exploiting the rotational symmetry of the $\theta$-$\phi$ subspace, we can,
without loss of generality, work in the equatorial plane, considering only
geodesics which have $\chi^\theta (\lambda) = \pi /2$.  In this case, Equation
\ref{normalization} becomes
\begin{equation}
    \label{eq:pr-normalization}
-\frac{e^2}{1 - \frac{2}{\chi^r}} + \frac{l^2}{(\chi^r)^2} + \frac{1}{1-\frac{2}{\chi^r}} \left( \frac{\dd \chi^r}{\dd \lambda} \right)^2 = 0.
\end{equation}
Rescaling the affine parameter, we obtain
\begin{equation}
\label{radial-path}
\frac{\dd \chi^r(s)}{\dd s} = \pm \sqrt{1 - \frac{b^2\left(1 - \frac{2}{\chi^r(s)} \right)}{(\chi^r(s))^2}},
\end{equation}
with $b \equiv l/e$ called the ``impact parameter''\footnote{$b$ is so named
because it gives the distance at infinity between the geodesic and the nearest
parallel line running radially from the center of the star.} of the trajectory.
This equation and Equations \ref{eq:conserved-e} and \ref{eq:conserved-l}
completely determine a trajectory given $l$ and $e$.

@@

@@

To convert this world-line tangent vector to the \emph{observer} coordinate
system, let the inclination of the neutron star spin be $\iota$ and the
obliquity (measered ccw from ``vertical'' in the observer frame) be $\omega$.
Then the neutron star spin three-vector has cartesian components in the observer
frame of 
\begin{equation}
    \vec{s} = \left( \cos \iota, -\sin \iota \sin \omega, \sin \iota \cos \omega \right).
\end{equation}
See the Figure \ref{fig:obs_cart}.
%%%
\begin{figure}
    \centering
        \includegraphics[width=0.9\columnwidth]{obs_cart_coord.png}
        \caption{The coordinates in the observer frame.}
        \label{fig:obs_cart}
    \end{figure}
%%%
If the incoming ray strikes the surface at a point with colatitude $\theta_0$
and longitude $\phi_0$, then the vector pointing to the impact point has
cartesian components in the observer frame of 
\begin{equation}
    \vec{r}_0 = \left( \cos \phi_0 \sin \theta_0, \sin\phi_0 \sin \theta_0, \cos \theta_0 \right).
\end{equation}
The instantaneous world-line tangent of the observer fixed to the NS surface at
the point of impact is given in cartesian coordinates by
\begin{equation}
    \label{eq:tangent_vel}
    \vec{t}_0 = \frac{2 \pi r_{NS}}{P} \frac{\partial \chi^t}{\partial \lambda} \vec{s} \times \vec{r}_0. 
\end{equation}

@@

>>>>

The metric has two killing vectors, $\partial_t$ and $\partial_\phi$, since it
is independent of the corresponding coordinates.  Write
\begin{equation}
\partial_t = \xi^\mu \partial_\mu, \qquad \partial_\phi = \zeta^\mu \partial_\mu;
\end{equation}
then the only non-zero component of $\xi^\mu$ is $\xi^t = 1$, and the only
non-zero component of $\zeta^\mu$ is $\zeta^\phi = 1$.  If $p^\mu$ is the
momentum associated with a geodesic,
\begin{equation}
    \label{eq:momentum_definition}
p^\mu = \frac{\dd \chi^\mu(\lambda)}{\dd \lambda},
\end{equation}
with $\lambda$ either proper time (for time-like geodesics) or an affine
parameter (null geodesics), and $\chi^\mu(\lambda)$ the coordinates of the
points on the geodesic, then N\"{o}ther's Theorem gives 
\begin{eqnarray}
\label{eq:conserved-e}
g_{\mu \nu} \xi^\mu p^\nu = - \left( 1 - \frac{2}{\chi^r} \right) p^t & = & \textnormal{const} \equiv - e,\\
\label{eq:conserved-l}
g_{\mu \nu} \zeta^\mu p^\nu = (\chi^r)^2 \sin^2 (\chi^\theta) p^\phi & = & \textnormal{const} \equiv L
\end{eqnarray}
along any geodesic.  ($e$ and $L$ are the energy and angular momentum---per unit
mass, if appropriate---of the particle traveling the geodesic measured by an
observer at infinity.)

For a null geodesic, we also have the normalization condition 
\begin{equation}
\label{normalization}
g_{\mu \nu} p^\mu p^\nu = 0.
\end{equation}
Exploiting the rotational symmetry of the $\theta$-$\phi$ subspace, we can,
without loss of generality, work in the equatorial plane, considering only
geodesics which have $\chi^\theta (\lambda) = \pi /2$.  In this case, Equation
\ref{normalization} becomes
\begin{equation}
    \label{eq:pr-normalization}
-\frac{e^2}{1 - \frac{2}{\chi^r}} + \frac{l^2}{(\chi^r)^2} + \frac{1}{1-\frac{2}{\chi^r}} \left( \frac{\dd \chi^r}{\dd \lambda} \right)^2 = 0.
\end{equation}
Rescaling the affine parameter, we obtain
\begin{equation}
\label{radial-path}
\frac{\dd \chi^r(s)}{\dd s} = \pm \sqrt{1 - \frac{b^2\left(1 - \frac{2}{\chi^r(s)} \right)}{(\chi^r(s))^2}},
\end{equation}
with $b \equiv L/e$ called the ``impact parameter''\footnote{$b$ is so named
because it gives the distance at infinity between the geodesic and the nearest
parallel line running radially from the center of the star.} of the trajectory.
This equation and Equations \ref{eq:conserved-e} and \ref{eq:conserved-l}
completely determine a trajectory given $L$ and $e$.

@@

From the Equation \ref{eq:conserved-l}, we can write
\begin{equation}
    p^{\phi} = \frac{L}{(\chi^r)^2 \sin^2(\chi^\theta)} = \frac{L}{(\rNS)^2 \sin^2(\cos^{-1} \left( \sin \alpha \sin \psi \right))}
\end{equation}

and after applying the normalization condition, we get
\begin{equation}
    p^{\theta} = -\frac{\sqrt{\frac{l^2\left(\frac{e^2\left(l^2\left(3-2\rNS^2\right)^2-4\rNS^2\left(\rNS^2-3\right)\right)}{\left(l^2-1\right)(\rNS-2)}+\frac{4\rNS}{\sin^2(\alpha)\sin^2(\psi)-1}\right)}{\rNS^3}}}{2\rNS}
\end{equation}

or, if we expand in series, then we get
\begin{equation}
    p^{\theta}=-\frac{\sqrt{e^2l^2}}{\rNS}-\frac{\sqrt{e^2l^2}}{\rNS^2}-\frac{\sqrt{e^2l^2}}{2\rNS^3\left(e^2\left(\sin^2(\alpha)\sin^2(\psi)-1\right)\right)}+O\left(\left(\frac{1}{\rNS}\right)^4\right)
\end{equation}

@@

To convert this world-line tangent vector to the \emph{observer} coordinate
system, let the inclination of the neutron star spin be $\iota$ and the
obliquity (measered ccw from ``vertical'' in the observer frame) be $\omega$.
Then the neutron star spin three-vector has cartesian components in the observer
frame of 
\begin{equation}
    \vec{s} = \left( \cos \iota, -\sin \iota \sin \omega, \sin \iota \cos \omega \right).
\end{equation}
See the Figure \ref{fig:obs_cart}.
%%%
\begin{figure}
    \centering
        \includegraphics[width=0.9\columnwidth]{obs_cart_coord.png}
        \caption{The coordinates in the observer frame.}
        \label{fig:obs_cart}
    \end{figure}
%%%
If the incoming ray strikes the surface at a point with colatitude $\theta_0$
and longitude $\phi_0$, then the vector pointing to the impact point has
cartesian components in the observer frame of 
\begin{equation}
    \vec{r}_0 = \left( \cos \phi_0 \sin \theta_0, \sin\phi_0 \sin \theta_0, \cos \theta_0 \right).
\end{equation}
The instantaneous world-line tangent of the observer fixed to the NS surface at
the point of impact is given in cartesian coordinates by
\begin{equation}
    \label{eq:tangent_vel}
    \vec{\tau}_0 = \frac{2 \pi r_{NS}}{P} \frac{\partial \chi^t}{\partial \lambda} \vec{s} \times \vec{r}_0. 
\end{equation}

@@

\section{Notations}
\begin{table}[]
    \begin{tabular}{lll}
    \hline
    \textbf{Symbol}                   & \textbf{Description}                                                                & \textbf{Reference}                                        \\ \hline
    $\theta_0$                        & colatitude on the surface of the star                                                &                                                         \\
    $\phi_0$                          & longitude on the surface of the star                                                 &                                                         \\
    $L$                              & the specific angular momentum                                                        & Equation \ref{eq:conserved-l}                           \\
    $t$                              & the clock time of an observer at $r=\infty$                                            & --                                                      \\
    $\rNS$                           & radius of a neutron star                                                             &                                                         \\
    $p^\mu$                          & the momentum associated with a geodesic                                              & Equation \ref{eq:momentum_definition}                   \\
    $\chi^\mu$                       & the coordinates of the points on the geodesic                                        & --                                                      \\
    $\lambda$                        & affine parameter                                                                     & --                                                      \\
    $g_{\mu\nu}$                     & components of the metric tensor                                                      & --                                                      \\
    $e$                              & the specific energy                                                                  & Equation \ref{eq:conserved-e}                           \\
    $\omega$                         & the stellar obliquity                                                                &                                                         \\
    $\iota$                          & the stellar inclination                                                             &                                                        \\
    $\psi$                           & the angular components of the impact parameter \\
                                     &  associated with the positive $x$-axis &                                            \\
                                     &                                                                                      &                                                         \\ \hline
    \end{tabular}
    \end{table}
